Sperm should evolve to make female meiosis fair.

### 2014-07-27 writeup_and_notes/GeneticsResub/male_dependent_drive.tex
yb (nearly) finalized changes for resub

{\bf Abstract:}
Genomic conflicts arise when an allele gains an evolutionary advantage at a cost to organismal fitness. 
O\"{o}genesis is inherently susceptible to such conflicts
because alleles compete \yb{for inclusion in} the  product of female meiosis transmitted to the egg. 
Alleles that distort meiosis in their favor (i.e. meiotic drivers) often decrease organismal fitness,
	and therefore indirectly favor the evolution of mechanisms to suppress meiotic drive. 
In this light, many facets of  o\"{o}genesis and gametogenesis have been 
	interpreted as mechanisms of protection against genomic outlaws. 
The fact that female meiosis \yb{is} often left uncompleted until after
      fertilization in many animals appears to run counter to this interpretation, 
      as this delay provides an opportunity for sperm\yb{-acting} alleles to
     meddle with its outcome and help like-alleles drive in heterozygous females. 
%Why then is female meiosis often left uncompleted until after
%fertilization in many animals -- 
%	potentially providing an opportunity for sperm alleles to
%     meddle with its outcome and help like-alleles drive in heterozygous females? 
The population genetic theory presented herein suggests that
	\gc{in fact} sperm nearly always evolve to increase the fairness of female
        meiosis in the face of genomic conflicts. 
These results are consistent with \yb{the apparent sperm- dependence}  \gc{of the best characterized female meiotic drivers in animals to date.} 
 \yb{Rather than providing an opportunity for sperm collaboration in female meiotic drive,
        the `fertilization requirement'  indirectly protects females from
        meiotic drivers by providing sperm an opportunity to
        suppress female meiotic drive. }
%	and suggest that the requirement of fertilization for the
%%        completion of female meiosis
%        potentially represents a mechanism employed by females 
%	 to ensure a fair meiosis.
\newpage

@@

It is therefore somewhat surprising that despite the intense evolutionary pressure on female meiosis to prevent meiotic drive, 
it is potentially open to sabotage by a virtual stranger -- a haploid sperm genome.
That is, in many animal species, female meiosis is completed only after fertilization \citep{Masui_book}, 
	creating ample opportunity for interaction between the sperm and
	female meiotic machinery (note that the alternation of generations
	likely precludes this interaction in plants). 
Therefore, a `green-bearded' \citep{Gardner2010} sperm-acting allele that recognizes and facilitates the meiotic drive of a genetically equivalent allele in heterozygous females %female meiosis %(a  heteromorphic dyad), 
     could presumably rapidly spread through a population. 
At first sight, female meiosis appears primed for \gc{conflict caused by
the such selfish systems.} Here we ask if sperm do indeed evolve to collaborate with female drivers to exploit this apparent weakness in the defense against meiotic drive. \newline

Before doing so, it is worth considering whether there is any evidence that sperm could (or do), influence female meiosis. 
It is becoming
        increasingly clear that sperm bring a wide variety of RNA and proteins into the
        egg \citep{Miller:05}. 
 Some of these have known functions, for example, in most
        animal species sperm-- not eggs-- are
        responsible for the transmission of the centriole, a vital
        component of the mitotic machinery for the zygote \citep{Schatten:94}. 
Detailed functional studies and paternal effect mutations in model
systems further highlight
 that sperm-transmitted products have a wide-range of functions in
 the \gc{egg activation, completion of syngamy, zygotic development, and the resumption
 and successful completion of
 female meiosis} \cite[e.g.][]{Yasuda:95,Loppin:05,Miller:01, McNally:05,Churchill:03}. 
 For example, in \emph{C. elegans}, \gc{premature deployment of the sperm aster 
	disrupts MII meiotic segregation in the egg, leading to a triploid zygote} \citep{McNally2012}.
However,
        the function of many of the products the sperm brings into the
        egg is completely unknown and 
      these products vary widely over species \citep{Karrbook:09}. 
It seems quite plausible that sperm-based products and
hence alleles could potentially have functional effects on
various aspects of female meiosis that occur after fertilization. 
%Why then is the requirement of fertilization to complete female meiosis so ubiquitous? 
\newline

%It is certainly not the case that sperm is mechanistically incapable of influencing female meiosis.
%Both mechanistic and genetic evidence suggests that sperm can influence the outcome of female meiosis.
%An example of mechanistic evidence for sperm's influence on female meiosis comes from \emph{C. elegans}, where premature deployment of the aster -- a vital component of mitotic machinery provided by the sperm, 
%	disrupts MII meiotic segregation in the egg, leading to a triploid zygote \citep{McNally2012}.
Current evidence suggests that sperm influence on female meiotic drive is not only possible, but likely. 
Evidence from both the \emph{In} and \emph{Om}  -- suggests that the 
	 transmission of alternate alleles in heterozygous females can depend on sperm haplotype.  
%There are current only two polymorphic, well-characterized female meiotic drive systems in
%animals-- and both of them, the \emph{In} and \emph{Om}  drive systems in mice-- suggests that the 
	 transmission of alternate alleles in heterozygous females can depend on sperm haplotype.  
 While ruling out the alternative hypothesis of early selection on zygotes in these cases is challenging  \citep[see pages 52-54 in][ for comment]{Burt2006}, it appears that the extent to which \emph{In} and \emph{Om} distort the second female meiotic division partially depends on the 	
	genotype of the fertilizing sperm \citep{Agulnik1993,Wu2005}. 
\yb{The fact that the two best characterized, polymorphic systems of putative female 
	meiotic drive systems identified to date in animals (\emph{In} and \emph{Om}) 
	show this effect suggests that if female meiotic drive is
	common the role of sperm in modifying female drive will be important as well. } \newline

\yb{Numerous lines of evidence suggest that female meiotic drive is an 
	evolutionary force, and therefore the opportunity for sperm to influence female drive likely had applications to many animals}.   
While, research to date has identified a few extreme cases of female meiotic 
	drive in the small number model systems systematically
        studied \citep{Agulnik1990, Fishman2008, Hiatt2003,Novitski1951, Pardo-ManuelDeVillena2001c},
	rapid evolution of centromeres, telomeres and other basic components of the meiotic machinery, 
	suggest consistent evolution of female meiotic drivers and suppressors of meiotic drive
	 \citep[e.g. ][]{Anderson2008, Anderson2009, Axelsson2010,
           Malik2009a}. 
 We expect that over the next decade the spread of sequencing to a
         range of systems will reveal many more female meiotic drive
         systems, however, carefully characterizing them will still
         remain a challenging task.
\newline

@@

\yb{Because female meiotic drive is likely common and consequential, 
	and because sperm have ample opportunity to influence female drive,  
	we develop population genetic models exploring the evolution of sperm influence on female meiotic drive -- 
	that is, what is the expected influence of sperm on female drive? } 
We first focus on models in which `self-promoting' alleles in
  sperm facilitate drive of like alleles during gametogenesis in heterozygous females. 
These models show that such sperm-acting alleles 
	have more difficulty invading a population than do traditional meiotic drivers, 
	 and under most circumstances, cannot be maintained as a balanced polymorphism.
Because self-promoting drivers are unlikely to create a sustained genomic conflict, 
	female meiosis will have little incentive to evolve resistance to them.
We then examine models in which a novel sperm-acting allele modifies the efficacy of a polymorphic meiotic driver. 
Such models universally reveal that sperm-acting drive modifiers are favored only if they suppress drive. 
These results highlight the fact that the interests of sperm and
maternal  genomes' are often aligned, 
as both are invested in the fate of the resultant zygote (as was speculated for the \emph{In} locus \citep{Pomiankowski1993}).
Thus, there is little selective benefit to females in preventing sperm to influence female meioses. 
\gc{Given the wide-spread requirement of fertilization for the completion
of  female meiosis it seems possible that various features of the
interaction between sperm and egg have evolved as a mechanism to enforce an equitable transfer of genetic material.}

@@

\yb{
\section*{Methods}
We present deterministic one- and two- locus population genetic models of sperm influence on female meiotic drive to evaluate wether sperm are likely to collaborate with female meiotic drivers or to stop them. \newline

We present six related models, the first three of which represent the system with a single locus.
{\bf{Model 1}} describes a traditional, single-locus, female meiotic driver, 
{\bf{Model 2}} describes a single-locus sperm-dependent female driver -- that is, an allele whose transmission in female meiosis depends on sperm haplotype. 
{\bf{Model 3}} describes a single-locus paternal-dependent female driver -- an allele whose transmission in female meiosis depends on paternal genotype. 
Assuming that a traditional driver segregates at its equilibrium frequency ({\bf{Model 1}}), we investigate the evolution of tightly linked ({\bf{Models 4 and 5}}), and unlinked ({\bf{Model 6}}) sperm-dependent modifiers of drive (We also briefly compare Model 6, to Model 6' - the suppression of female drive by an unlinked female acting suppressor).  
In {\bf{Models 4 and 5}}, we treat this two-locus system as if it consists of
a single locus with three alleles: $A$, $B$ and $C$, corresponding to the case when the
sperm-modifier arises once, tightly linked to one or other of the
alleles at the drive locus (such that recombination is unexpected). 
In all cases, we assume that fitness is independent of the drive modifier. 
 \newline

All models include a biallelic locus ($A/B$) with non-driving and driving alleles in frequencies $f_A$ and $f_B$, 
	respectively, while more complex models include more loci and/or alleles.  
Transmission rules describing the outcomes of all matings in each
	model are presented in a File S1 . 
The fitness of genotype, $g$, is sex-independent and equals $1$, $1-h_s$, and $1-s$ for genotypes $AA$, $AB$, and $BB$, respectively. %  and  $w_{AA} = 1$, $w_{AB} = 1-s_h$, and $w_{BB}=1-s$. 
Genotypic frequencies  equal  $f_g$ for adults in the current generation, 
	$f_g'$ in the next generation of zygotes (i.e. after recombination, random mating, drive, and syngamy, but before selection)
	and $f_g''$ in adults after selection. 
After a complete generation $f_g'' = f_g' w_g/\bar{w}$, where $\bar{w}$ is the population mean fitness and equals $\Sigma w_g f_g'$.  \newline

We verbally describe our main results below.  
Readers interested in the details of these results should turn to the Appendix for a mathematical treatment. 
There we present critical analytical results in Equations 1-11, and describe our analyses and results in more detail.  
Because a number of our analyses are approximations based on assuming
that genotype frequencies follow Hardy Weinberg Equilibrium (HWE), we note which analytical results are approximations. 
We verify these approximations with exact numerical iterations in Figures S1-S3 and File S3 , and detail the derivation of these approximations in our File S2.
}

@@

In contrast to a traditional driver, which drives but pays effectively
	no fitness cost when rare, 
a self-promoting driver specifically creates low fitness drive homozygotes 
	by uniting driving female gametes with sperm enabling that drive.
It must therefore overcome a drive-associated homozygous fitness cost simply to spread when rare. 
The conditions allowing the invasion of a self-promoting driver
 	are consequently far more restrictive than those for a standard meiotic driver.
When rare, a fully recessive, self-promoting driver can only invade when $s$ 
	is less than approximately $(2 d - 1)/(4 d)$ -- see dashed black line in Figure \ref{Eggsperm}B. 
This analytical approximation, derived from Equation \eqref{deltadriver} assuming Hardy-Weinberg, 
	closely matches results obtained by exact numerical iteration (Figure \ref{Eggsperm}B. \yb{We note that Equation 1 and all equations discussed in the main text are presented in the appendix and derived in File S2}). \newline

@@

%%%sperm-acting drive suppressors
While  sperm enhancement of a female drive cannot displace a polymorphic female driver, sperm based drive suppressors can. 
Imagine a sperm-acting allele that restores fairness to female meiosis arises on
	the drive background, creating a
        third allele $B^{-}$ (Figure
          \ref{Eggsperm_3_allele_cartoon}A3, Model 4 in Appendix). This new allele still
        experiences female drive when fertilized by A or B sperm, but 
       it does not drive when fertilized by another $B^{-}$ so it
       avoid the excess formation of low fitness homozygotes. This
     allows the $B^{-}$ to displace the ancestral driver (Figure
     \ref{Eggsperm_3_allele_cartoon}B1, Equation \ref{coupling}), 
	and often returns to a lower equilibrium frequency than the the $B$ allele, further decreasing the extent of drive in the population.  
If this sperm-acting drive suppressor arises on 
	the non-driving $A$ background ( i.e. in repulsion phase, 
	creating a third allele $A^{-}$, Figure \ref{Eggsperm_3_allele_cartoon}A4, Model 5), 
	or is unlinked to the drive locus (Model 6), it readily invades a population segregating
	for the drive system (Equations \ref{A-} and
        \ref{unlinked}).  %%DELETED FIG REF HERE. 
        \newline

The sperm acting drive suppressing allele lowers the frequency of the original driver (perhaps to zero),
	and spreads to fixation if it does not carry strong fitness costs
	(Figure \ref{Eggsperm_3_allele_cartoon}B2). 
This result is consistent with previous work showing that drive suppressors unlinked to, or in repulsion phase with drivers usually invade polymorphic drive systems \citep[e.g. ][]{Brandvain2012}.  
\yb{We note that the evolution of sperm-acting drive  suppressors unlinked to a driver (Model 6) is both qualitatively and quantitatively similar to the evolution  of a female-acting drive suppressor (compare Equations \ref{unlinked} and \ref{femunlinked}).}
Therefore, all two-locus models of sperm influence on female drive suggest that
	sperm will evolve to oppose female meiotic drive. 
\newline

@@

It seems that allowing sperm to influence the outcome of female meiosis would generate a confluence of these potential conflicts -- 
	sperm could actually assist an allele that distorts female meiosis.
However, this is not the case.
We find that an allele which acts through sperm to distort female meiosis in its favor %YB [too awkward / unclear?] 
	can rarely spread through a population if it bears any cost. 
Additionally, when this self-promoting driver can spread, it can only rarely 
	be maintained as a protected polymorphism, and due to its positive frequency dependence,  
	it spends very little time at intermediate frequency.
As such, this type of exploitation cannot generate a sustained genetic conflict.
It is therefore unlikely
	that female o\"{o}genesis and meiosis will evolve to prevent their effect.  
\gc{Thus females can} delay the completion of meiosis until after fertilization 
	without risking exploitation by collaborations between female
        drivers and sperm alleles. \newline 
%%G: felt like this made it sound less why they do it.

@@

\yb{The theory presented herein provides a strong prediction -- 
organisms in which female 
meiosis is not completed until after fertilization sperm will act to suppress female drive at the stage at which they can influence meiosis. 
%Therefore, for example, when fertilized by drying sperm, we expect meiosis to be fairer when  females are fertilized by 
Contrary to expectations of a \gc{naive verbal} ``green-beard'' model, 
we are unsurprised by the observation that female meiosis is fairer when fertilized by sperm bearing the drive allele in two of the best described cases of female meiotic drive in animals (the \emph{Om} and \emph{In} loci in mice \citep{Agulnik1993,Wu2005}).  
Given the immense benefit to sperm of hampering female drive, 
drive-suppressing sperm are often likely to be fixed within a species, 
making the hypothesis of sperm acting drive-suppression difficult to test from intra-population crosses. 
However, crosses between populations or species are likely to provide critical tests of our theory -- specifically we predict that female meiosis will be less fair when a species (or population) is fertilized by heterospecific sperm because such sperm have not evolved to counter novel females meiotic drivers. 
We further predict that segregation in F1 females backcrossed to parental species will likely be biased with a deficit of transmission of the paternal species allele from the F1  female.   
We note that while these predictions follow straightforwardly from the
theory presented above, tests of meiotic drive, 
and especially sperm dependent meiotic drive require a high standard
of evidence 
to exclude plausible alternative hypotheses such as genotypic inviability including epistatic maternal X zygotic lethality.}  \newline

\gc{Our theory also encourages phylogenetic hypotheses concerning the relationship between the opportunity 
	for female meiotic drive and the requirement of fertilization for the completion of female meiosis. 
For example, we predict that a lower opportunity for female meiotic drive,
        e.g. an animal lineage with a history of high inbreeding
        or selfing, 
        may be accompanied by a relaxation of the requirement of fertilization for the
        completion of female meiosis (although 
        opportunities to test this hypothesis may be limited because 
        %this effect may be weaks as such
        lineages may only persist for a short time). 
Our results also suggests that 
%Another   hypothesis that our work suggests is we might expect 
        that phylogenetic variation in the stage of female meiosis when
        fertilization occurs may influence the prevalence of
        female meiotic drive.
\gc{For example, in taxa where meiosis I is completed
by the egg pre-fertilization centromere drive may
be more common as sperm-based modifiers can not intercede, as compared to
species where fertilization occurs during MI. }
Such hypotheses could be tested if the speed of centromere
turnover serves as a proxy for the efficacy of MI drive. 
\newline

@@

\subsubsection*{Model $6^\prime$. Unlinked female acting drive-modifier:}
\yb{The dynamics of a female-acting drive modifier are comparable to those describing a female acting drive suppressor.  
When rare and fully dominant and in linkage and in HWE, the change in frequency of a rare unlinked, female-acting drive modifier is 
\begin{equation} \Delta M =-(d_h-d_0) f_A f_B (f_B s + s_h (f_A - f_B) ) \frac{f_M}{\bar{w}}  \label{femunlinked} \end{equation}
When  when drive-modification is dominant ($d_h = d_1 = d_0 + \epsilon$), Equation \ref{femunlinked} is equal to Equation \ref{unlinked}.
 However, if female drive suppression is less than fully dominant, sperm-acting drive suppressors are more efficacious when rare than are female-acting suppressors, and are therefore more likely to spread.  
}

>>>>

{\bf Abstract:}
Genomic conflicts arise when an allele gains an evolutionary advantage at a cost to organismal fitness. 
O\"{o}genesis is inherently susceptible to such conflicts
because alleles compete for inclusion in the  product of female meiosis that will be transmitted to the egg. 
Alleles that distort meiosis in their favor (i.e. meiotic drivers) often decrease organismal fitness,
	and therefore indirectly favor the evolution of mechanisms to suppress meiotic drive. 
In this light, many facets of  o\"{o}genesis and gametogenesis have been 
	interpreted as mechanisms of protection against genomic outlaws. 
That females  of many animal species do not complete meiosis until after
     fertilization appears to run counter to this interpretation
     because this delay provides an opportunity for sperm-acting alleles to
     meddle with the outcome of female meiosis and help like-alleles drive in heterozygous females. 
Contrary to this perceived danger, the population genetic theory presented herein suggests that, 
	in fact, sperm nearly always evolve to increase the fairness of female
        meiosis in the face of genomic conflicts. 
These results are consistent with the apparent sperm-dependence of the best characterized female meiotic drivers in animals to date.  
Rather than providing an opportunity for sperm collaboration in female meiotic drive,
        the `fertilization requirement'  indirectly protects females from
        meiotic drivers by providing sperm an opportunity to
        suppress drive. 
%	and suggest that the requirement of fertilization for the
%%        completion of female meiosis
%        potentially represents a mechanism employed by females 
%	 to ensure a fair meiosis.
\newpage

@@

It is therefore somewhat surprising that despite the intense evolutionary pressure on female meiosis to prevent meiotic drive, 
it is potentially open to sabotage by a virtual stranger -- a haploid sperm genome.
That is, in many animal species, female meiosis is completed only after fertilization \citep{Masui_book}, 
	creating ample opportunity for interaction between the sperm and
	female meiotic machinery (note that the alternation of generations
	likely precludes this interaction in plants). 
Therefore, a `green-bearded' \citep{Gardner2010} sperm-acting allele that recognizes and facilitates the meiotic drive of a genetically equivalent allele in heterozygous females %female meiosis %(a  heteromorphic dyad), 
     could presumably rapidly spread through a population. 
At first sight, female meiosis appears primed for conflict caused by such selfish systems. 
Here we ask if sperm do indeed evolve to collaborate with female drivers to exploit this 
	apparent weakness in the defense against meiotic drive. \newline

Before doing so, it is worth considering whether there is any evidence that sperm could (or do), influence female meiosis. 
It is becoming
        increasingly clear that sperm bring a wide variety of RNA and proteins into the
        egg \citep{Miller:05}. 
 Some of these have known functions, for example, in most
        animal species sperm -- not eggs -- are
        responsible for the transmission of the centriole, a vital
        component of the mitotic machinery for the zygote \citep{Schatten:94}. 
Detailed functional studies and analyses of paternal effect mutations in model
systems further highlight that sperm-transmitted products have a wide-range of functions in
 egg activation, completion of syngamy, zygotic development, and the resumption
 and successful completion of
 female meiosis \cite[e.g.][]{Yasuda:95,Loppin:05,Miller:01, McNally:05,Churchill:03}. 
 For example, in \emph{C. elegans}, premature deployment of the sperm aster 
	disrupts MII meiotic segregation in the egg, leading to a triploid zygote \citep{McNally2012}.
However, the function of many of the products the sperm brings into the
        egg is completely unknown and 
      these products vary widely over species \citep{Karrbook:09}. 
It seems quite plausible that sperm-based products, and
	hence sperm haplotype or paternal genotype could influence 
	various aspects of female meiosis that occur after fertilization. 
\newline

Current evidence from the best characterized systems of female meiotic drive in animals 
	(the  \emph{In} and \emph{Om} loci in mice) 
	suggests that sperm influence on female meiotic drive is not only possible, but likely. 
%Evidence from both the \emph{In} and \emph{Om}  -- suggests that the 
%	 transmission of alternate alleles in heterozygous females can depend on sperm haplotype.  
%There are current only two polymorphic, well-characterized female meiotic drive systems in
%animals-- and both of them, the \emph{In} and \emph{Om}  drive systems in mice-- suggests that the 
%	 transmission of alternate alleles in heterozygous females can depend on sperm haplotype.  
 While ruling out the alternative hypothesis of early selection on zygotes in these cases is challenging  \citep[see pages 52-54 in][ for comment]{Burt2006}, it appears that the extent to which \emph{In} and \emph{Om} distort the second female meiotic division partially depends on the 	
	genotype of the fertilizing sperm \citep{Agulnik1993,Wu2005}. 
The fact that the two best characterized, polymorphic systems of putative female 
	meiotic drive systems in animals 
	show this effect suggests that if female meiotic drive is common 
	the role of sperm in modifying female drive will be important as well.  \newline

Numerous lines of evidence suggest that female meiotic drive is a common and important 
	evolutionary force, and therefore the opportunity for sperm to influence female drive likely had applications to many animals.   
While research to date has identified a few extreme cases of female meiotic 
	drive in the small number model systems systematically
        studied \citep{Agulnik1990, Fishman2008, Hiatt2003,Novitski1951, Pardo-ManuelDeVillena2001c},
	rapid evolution of the basic components of the meiotic machinery (e.g. centromeres, telomeres, etc \dots) 
	suggest consistent selection on female meiotic drivers and suppressors of meiotic drive in many animal species 
	 \citep[e.g. ][]{Anderson2008, Anderson2009, Axelsson2010, Malik2009a}. 
 We expect that over the next decade the spread of sequencing to a
         range of systems will reveal many more female meiotic drive
         systems, however, carefully characterizing them will still
         remain a challenging task. 
\newline

@@

Because female meiotic drive is likely a common force with predictably negative effects on organismal fitness,  
	and because sperm have ample opportunity to influence female drive,  
	we develop population genetic models to address the the expected influence of sperm on female drive.  
We first focus on models in which `self-promoting' alleles in
  sperm facilitate drive of like alleles during gametogenesis in heterozygous females. 
These models show that such sperm-acting alleles 
	have more difficulty invading a population than do traditional meiotic drivers, 
	 and under most circumstances, cannot be maintained as a balanced polymorphism.
Because self-promoting drivers are unlikely to create a sustained genomic conflict, 
	female meiosis will have little incentive to evolve resistance to them.
We then examine models in which a novel sperm-acting allele modifies the efficacy of a polymorphic meiotic driver. 
Such models universally reveal that sperm-acting drive modifiers are favored only if they suppress drive. 
These results highlight the fact that the interests of sperm and
maternal  genomes' are often aligned, 
as both are invested in the fate of the resultant zygote (as was speculated for the \emph{In} locus \citep{Pomiankowski1993}).
Thus, there is little selective benefit to females in preventing sperm to influence female meioses, 
	\yb{and in fact, females eschewing this delay would potentially lose sperm assistance in the suppression of meiotic drivers.}  
Given the wide-spread requirement of fertilization for the completion
	of  female meiosis, various features of the interaction between sperm 
	and egg likely evolved as a mechanism 
	to enforce an equitable transfer of genetic material.

@@

\section*{Methods}
We present deterministic one- and two- locus population genetic models of sperm influence on female meiotic drive to evaluate wether sperm are likely to collaborate with female meiotic drivers or to stop them. \newline

We present six related models -- three single-locus `pleiotropy' models and three two-locus `drive-modifier' models.
	{\bf{Model 1}} describes a single-locus female meiotic driver.  
	{\bf{Model 2}} describes a single-locus sperm-dependent female driver -- that is, an allele whose transmission in female meiosis depends on sperm haplotype. 
	{\bf{Model 3}} describes a single-locus paternal-dependent female driver -- an allele whose transmission in female meiosis depends on paternal genotype. 
Assuming that a traditional driver segregates at its equilibrium frequency (identified in Model 1), 
	we investigate the evolution of tightly linked ({\bf{Models 4 and 5}}), and unlinked ({\bf{Model 6}}) 
	sperm-dependent modifiers of drive. 
In {\bf{Models 4 and 5}}, we treat this two-locus system as if it consists of
	a single locus with three alleles: $A$, $B$ and $C$, corresponding to the case when the
	sperm-modifier is very tightly linked to the driving ({\bf{Model 4}}) or non-driving allele ({\bf{Model 5}})
 	at the drive locus such that recombination is unexpected. 
To evaluate the feasibility of sperm-modification of female meiotic drive, as compared to female suppression of drive, 
	we conclude with a model of female drive suppression by an unlinked female acting suppressor ({\bf{Model $6^\prime$}}). 
In all cases, we assume that fitness is independent of the drive modifier. 
 \newline

All models include a biallelic locus ($A/B$) with non-driving and driving alleles in frequencies $f_A$ and $f_B = 1 - f_A$, 
	respectively, while Models 4-6 include a drive-modifying locus.  
Transmission rules describing the outcomes of all matings in each model are presented in a File S1. 
The fitness of genotype, $g$, is sex-independent and equals $1$, $1-h_s$, and $1-s$ for genotypes $AA$, $AB$, and $BB$, respectively. %  and  $w_{AA} = 1$, $w_{AB} = 1-s_h$, and $w_{BB}=1-s$. 
Genotypic frequencies  equal  $f_g$ for adults in the current generation, 
	$f_g'$ in the next generation of zygotes (i.e. after recombination, random mating, drive, and syngamy, but before selection)
	and $f_g''$ in adults after selection. 
After a complete generation genotype frequencies are $f_g'' = f_g' w_g/\bar{w}$, where $\bar{w}$ is the population mean fitness and equals $\Sigma w_g f_g'$.  \newline

We verbally describe our main results below.  
Readers interested in the details of these results should turn to the Appendix for a mathematical treatment, 
	and to our Mathematica worksheet  (File S2) for our complete derivations. 
There we present critical analytical results in Equations 1-11, and describe our analyses and results in more detail.  
Because a number of our analyses are approximations based on assuming
that genotype frequencies follow Hardy Weinberg Equilibrium (HWE), we note which analytical results are approximations. 
We verify these approximations with exact numerical iterations in Figures S1-S3 and File S3.

@@

In contrast to a traditional driver, which drives but pays effectively
	no fitness cost when rare, 
a self-promoting driver specifically creates low fitness drive homozygotes 
	by uniting driving female gametes with sperm enabling that drive.
It must therefore overcome a drive-associated homozygous fitness cost simply to spread when rare. 
The conditions allowing the invasion of a self-promoting driver
 	are consequently far more restrictive than those for a standard meiotic driver.
When rare, a fully recessive, self-promoting driver can only invade when $s$ 
	is less than approximately $(2 d - 1)/(4 d)$ -- see dashed black line in Figure \ref{Eggsperm}B. 
This analytical approximation, derived from Equation \eqref{deltadriver} assuming Hardy-Weinberg, 
	closely matches results obtained by exact numerical iteration (Figure \ref{Eggsperm}B. We remind readers that Equation 1 and all equations discussed in the main text are presented in the Appendix and derived in File S2). \newline

@@

%%%sperm-acting drive suppressors
While  sperm enhancement of a female drive cannot displace a polymorphic female driver, sperm based drive suppressors can. 
Imagine a sperm-acting allele that restores fairness to female meiosis arises on
	the drive background, creating a
        third allele $B^{-}$ (Figure
          \ref{Eggsperm_3_allele_cartoon}A3, Model 4 in Appendix). This new allele still
        experiences female drive when fertilized by A or B sperm, but 
       it does not drive when fertilized by another $B^{-}$ so it
       avoid the excess formation of low fitness homozygotes. This
     allows the $B^{-}$ to displace the ancestral driver (Figure
     \ref{Eggsperm_3_allele_cartoon}B1, Equation \ref{coupling}), 
	and often returns to a lower equilibrium frequency than the the $B$ allele, further decreasing the extent of drive in the population.  
If this sperm-acting drive suppressor arises on 
	the non-driving $A$ background ( i.e. in repulsion phase, 
	creating a third allele $A^{-}$, Figure \ref{Eggsperm_3_allele_cartoon}A4, Model 5), 
	or is unlinked to the drive locus (Model 6), it readily invades a population segregating
	for the drive system (Equations \ref{A-} and
        \ref{unlinked}).  %%DELETED FIG REF HERE. 
 We note that the evolution of sperm-acting drive  suppressors unlinked to a driver (Model 6)
 	is both qualitatively and quantitatively similar to the evolution  of a female-acting drive 
	suppressor (Model 6$^\prime$ -- compare Equations \ref{unlinked} and \ref{femunlinked}). 
        \newline

The sperm acting drive suppressing allele lowers the frequency of the original driver (perhaps to zero),
	and spreads to fixation if it does not carry strong fitness costs
	(Figure \ref{Eggsperm_3_allele_cartoon}B2). 
This result is consistent with previous work showing that drive suppressors unlinked to, or in repulsion phase with drivers usually invade polymorphic drive systems \citep[e.g. ][]{Brandvain2012}.  
Therefore, all two-locus models of sperm influence on female drive suggest that
	sperm will evolve to oppose female meiotic drive. 
\newline

@@

It seems that allowing sperm to influence the outcome of female meiosis would generate a confluence of these potential conflicts -- 
	sperm could actually assist an allele that distorts female meiosis.
However, this is not the case.
We find that an allele which acts through sperm to distort female meiosis in its favor %YB [too awkward / unclear?] 
	can rarely spread through a population if it bears any cost. 
Additionally, when this self-promoting driver can spread, it can only rarely 
	be maintained as a protected polymorphism, and due to its positive frequency dependence,  
	it spends very little time at intermediate frequency.
As such, this type of exploitation cannot generate a sustained genetic conflict.
It is therefore unlikely
	that female o\"{o}genesis and meiosis will evolve to prevent their effect.  
Thus, females can delay the completion of meiosis until after fertilization 
	without risking exploitation by collaborations between female
        drivers and sperm alleles. \newline 
%%G: felt like this made it sound less why they do it.

@@

The theory presented herein provides a strong prediction -- 
	for organisms in which female  meiosis is not completed until after fertilization, 
	sperm will act to suppress female drive at the stage at which they can influence meiosis. 
%Therefore, for example, when fertilized by drying sperm, we expect meiosis to be fairer when  females are fertilized by 
Contrary to expectations of a naive verbal ``green-beard'' model, 
	our results are consistent with the observation that female meiosis 
	is more fair when fertilized by sperm bearing the drive allele in 
	two of the best described cases of female meiotic drive in 
	animals (the \emph{Om} and \emph{In} loci in mice \citep{Agulnik1993,Wu2005}).  
Given the benefit to sperm of hampering female drive, 
	drive-suppressing sperm are often likely to be fixed within a species, 
	making the hypothesis of sperm-acting drive-suppression difficult to test from intra-population crosses. 
However, crosses between populations or species are likely to provide critical tests of our theory -- 
	specifically we predict that female meiosis will be less fair when a species (or population) is 
	fertilized by heterospecific sperm because such sperm have not evolved to counter novel females meiotic drivers. 
We further predict that segregation in F1 females backcrossed to parental species will likely be biased with a deficit 
	of transmission of the paternal species allele from the F1  female.   
We note that while these predictions follow straightforwardly from the
	theory presented above, tests of meiotic drive, 
	and especially sperm-dependent meiotic drive require a high standard
	of evidence  to exclude plausible alternative hypotheses such as genotypic 
	inviability including epistatic maternal X zygotic lethality \citep[e.g. ][]{Sawamura:1993aa}.  \newline

\gc{Our theory also encourages phylogenetic hypotheses concerning the relationship between the opportunity 
	for female meiotic drive and the requirement of fertilization for the completion of female meiosis. 
For example, we predict that a lower opportunity for female meiotic drive,
        e.g. an animal lineage with a history of high inbreeding
        or selfing, 
        may be accompanied by a relaxation of the requirement of fertilization for the
        completion of female meiosis (although 
        opportunities to test this hypothesis may be limited because 
        %this effect may be weaks as such
        lineages may only persist for a short time). 
Our results also suggests that 
%Another   hypothesis that our work suggests is we might expect 
        that phylogenetic variation in the stage of female meiosis when
        fertilization occurs may influence the prevalence of
        female meiotic drive.
For example, in taxa where females complete meiosis
	before fertilization, centromeric drive may
	be more common than in species which sperm interact with eggs arrested  in MI, 
	because in the prior case sperm-based modifiers can only intercede during the $2^\text{nd}$, but not $1^\text{st}$ meiotic division. }
As a potential test of this hypothesis the speed of centromere
	turnover could be compared in species in which sperm iterate with eggs paused at MI and MII 
	(assuming the pace of centromere turnover serves as a proxy for the efficacy of MI drive). 
\newline

@@

\paragraph{Model $6^\prime$. Unlinked female acting drive-modifier:}
The dynamics of a female-acting drive modifier are comparable to those describing a sperm-acting drive modifier.  
Assuming Hardy-Weinberg and linkage equilibrium, the change in frequency of a rare, unlinked, female-acting drive modifier is 
\begin{equation} \Delta M =-(d_h-d_0) f_A f_B (f_B s + s_h (f_A - f_B) ) \frac{f_M}{\bar{w}}  \label{femunlinked} \end{equation}
When  when drive-modification is dominant ($d_h = d_1 = d_0 + \epsilon$), Equation \ref{femunlinked} is equal to Equation \ref{unlinked}.
 However, if female drive suppression is less than fully dominant, sperm-acting drive suppressors are more efficacious when rare than are female-acting suppressors, and are therefore more likely to spread.
